Suitability prevalence area index in late quaternary explains genetic diversity in Tassel eared Squirrels

### 2023-07-11 manuscript_english.Rmd
expand introduction

>>>>

Relating the genetic structure of the populations of a species to its geographic distribution is a complicated challenge from the point of view of biogeography. The task is made easier if evolutionary history is incorporated on both sides of the relationship. 
On the one hand, giving depth and an evolutionary perspective to the geographic distribution of a taxon relies primarily on ecological niche theory, and is driven by the massive amount of information currently available from both digitized biological collections and citizen science (through which records of taxa occurrence are
through which records of taxa occurrence are available) as well as from environmental information available for geographic information systems and climate models. information systems and accurate climatological models of both the present and the past.

This change in reproductive success is driven by a set of factors, however we propose in this work a conceptual framework where we consider only environmental factors at a macroecological scale, inspired by the concept of Grinnellian niches (those composed by sets of points of non-interactive variables in multidimensional spaces), seeking as a principle the necessary conditions for fitness and not the response of taxa. points of non-interactive variables in multidimensional spaces), seeking as a principle the conditions necessary for fitness and not the response of taxa.

That is, we simplify our model (as is done in the physical sciences) to link through fitness the changes in biological populations and their structure, observed through population genetics, with patterns of spatial distribution over time. Thus, the goal is to link biogeographic patterns and evolutionary processes of populations in a single conceptual framework.

It is from this Grinellian niche where we start, assuming that taxa in general have to be distributed in the geography where environmental conditions are more favorable, noting that the distribution of taxa is not static, depending on dynamic environmental conditions and therefore the current known distribution of a taxon does not always correspond to its historical distribution over time, since the current climate has not remained so for periods of time on scales relative to the processes of population genetics. (i. e. tens or hundreds of thousands of years). Even for some species these dynamics seem to induce ecological processes such as seasonal migrations in birds, or hibernation in mammals and frogs. But in general, environmental changes are reflected in geography as range expansions and contractions over time. That is, processes of population dynamics, driven by environmental fitness, in our approach.

This change in population dynamics is implied by the genetic structure of populations through the concept of effective population size $N_e$ [@wright1931evolution; @karlin1968rates]. 
So if we set a similar time scale for these changes, we will be able to observe the fitness relationship (through population dynamics) in both genetic structure and distribution range changes over time.  That is, we need an index that reflects environmental changes in the geographic distribution of species and changes in the genetic structure of the population. In this case, we can consider the fixation index $F_{st}$ as an indicator of the difference between populations that come from a source population and that have undergone range shifts that we assume are driven by environmental changes. Under these assumptions we expect that it is possible to relate changes in the distribution of taxa to the genetic structure of their populations using statistical models to explain the relationship.

### 2023-07-11 manuscript_english.tex
expand introduction

@@

\leavevmode\vadjust pre{\hypertarget{ref-wright1943isolation}{}}%
Wright, Sewall. 1943. {``Isolation by Distance.''} \emph{Genetics} 28
(2): 114.

>>>>

Relating the genetic structure of the populations of a species to its
geographic distribution is a complicated challenge from the point of
view of biogeography. The task is made easier if evolutionary history is
incorporated on both sides of the relationship. On the one hand, giving
depth and an evolutionary perspective to the geographic distribution of
a taxon relies primarily on ecological niche theory, and is driven by
the massive amount of information currently available from both
digitized biological collections and citizen science (through which
records of taxa occurrence are through which records of taxa occurrence
are available) as well as from environmental information available for
geographic information systems and climate models. information systems
and accurate climatological models of both the present and the past.

This change in reproductive success is driven by a set of factors,
however we propose in this work a conceptual framework where we consider
only environmental factors at a macroecological scale, inspired by the
concept of Grinnellian niches (those composed by sets of points of
non-interactive variables in multidimensional spaces), seeking as a
principle the necessary conditions for fitness and not the response of
taxa. points of non-interactive variables in multidimensional spaces),
seeking as a principle the conditions necessary for fitness and not the
response of taxa.

That is, we simplify our model (as is done in the physical sciences) to
link through fitness the changes in biological populations and their
structure, observed through population genetics, with patterns of
spatial distribution over time. Thus, the goal is to link biogeographic
patterns and evolutionary processes of populations in a single
conceptual framework.

It is from this Grinellian niche where we start, assuming that taxa in
general have to be distributed in the geography where environmental
conditions are more favorable, noting that the distribution of taxa is
not static, depending on dynamic environmental conditions and therefore
the current known distribution of a taxon does not always correspond to
its historical distribution over time, since the current climate has not
remained so for periods of time on scales relative to the processes of
population genetics. (i. e. tens or hundreds of thousands of years).
Even for some species these dynamics seem to induce ecological processes
such as seasonal migrations in birds, or hibernation in mammals and
frogs. But in general, environmental changes are reflected in geography
as range expansions and contractions over time. That is, processes of
population dynamics, driven by environmental fitness, in our approach.

This change in population dynamics is implied by the genetic structure
of populations through the concept of effective population size \(N_e\)
(Wright 1931; Karlin 1968). So if we set a similar time scale for these
changes, we will be able to observe the fitness relationship (through
population dynamics) in both genetic structure and distribution range
changes over time. That is, we need an index that reflects environmental
changes in the geographic distribution of species and changes in the
genetic structure of the population. In this case, we can consider the
fixation index \(F_{st}\) as an indicator of the difference between
populations that come from a source population and that have undergone
range shifts that we assume are driven by environmental changes. Under
these assumptions we expect that it is possible to relate changes in the
distribution of taxa to the genetic structure of their populations using
statistical models to explain the relationship.

@@

\leavevmode\vadjust pre{\hypertarget{ref-wright1931evolution}{}}%
Wright, Sewall. 1931. {``Evolution in Mendelian Populations.''}
\emph{Genetics} 16 (2): 97.

\leavevmode\vadjust pre{\hypertarget{ref-wright1943isolation}{}}%
---------. 1943. {``Isolation by Distance.''} \emph{Genetics} 28 (2):
114.
